Explainable AI with Probabilistic Graphical Models

### 2019-11-13 introduction/response.tex

\section{Response} \label{sec:response}
In order to contextualise the current work, Chapter \ref{chap:literature-review} will investigate the notion of \textit{explainability}, its importance and how to evaluate it, as defined in the current xAI literature.
In particular, the explainability of Bayesian networks will be reviewed in detail as this model will be the basis for the methods carried out in this thesis.

To address the gaps identified in the previous section and in the following chapter, the work conducted in this thesis will concentrate on explainability in the medical domain and will present both a practical and a theoretical part.
The methods will include the implementation of a Bayesian network-based system, inspired by the work by \citet{Butz2018} (see Section \ref{sec:explaining-the-most-probable-explanation}), and its subsequent evaluation.
This recent xAI paper never provided any results for the compelling methods it presented, thus a proof of concept system implementing them, together with novel extensions, will be developed.
The system will be a means of exploring the efficacy of the explanatory modes for BNs - as identified by \citet{lacave2002review} - mainly the \textit{graphical}, \textit{linguistic} and, in particular, the \textit{dialogical}.
Their taxonomy for BN explanations, together with the psychological characteristics of an explanation \citep{miller2018explanation}, will be the framework against which the methods developed in Chapters \ref{chap:methodology} and evaluated in Chapter \ref{chap:results} will be measured.
The explainability of the work will be validated by real medical experts, in a concrete setting, over a period of time; this will also be a means to provide a validation for the methods of the initial paper \citep{Butz2018} and, more in general, of Bayesian networks as a whole.
The system will be a proxy to explore the explainability of BNs that should, in theory, be well adapted to giving highly effective explanations (as discussed in Section \ref{sec:explainability-in-bayesian-networks}).
The hope is also to set a methodological precedent for other \textit{application-grounded evaluations} (see \citep{doshi2017towards} and Section \ref{sec:evaluation-of-explainability}), with the aim of helping to reduce the gap in the xAI literature of the absence of actual human evaluation of explainability.

The methodology that will be used to evaluate these hypotheses will start with the study of a specific, recent xAI paper by \citet{Butz2018}, whose method will become the basis for the work carried out.
Then a proof of concept system will be developed, coded in Python and that will implement the learning and updating of a Bayesian network (see Subsections \ref{subsec:learning-bn-structure}, \ref{subsec:learning-bn-parameters} and \ref{subsec:bnupdating}) together with standard methods (see Subsection \ref{subsec:algorithms}) and novel algorithms (see Subsection \ref{subsec:algorithms-novel}).
The system's main aim will be to \textit{support medical decision making} by establishing a dialogue with the domain expert user and to evaluate the usefulness, in terms of explainability, of various other interaction modes such as conditional probability and most probable explanation queries (see Subsection \ref{subsec:bnupdating}); various algorithms and interaction paradigms will be developed to do so, as detailed in Section \ref{sec:novel-contributions}.
Finally, this system will be evaluated by expert pathologists at \textit{Istituto Cantonale di Patologia} of Locarno (Switzerland), the partner institution who will collaborate on the testing and evaluation of the software tool that will be developed.
The data set that will be used to learn the Bayesian network, composed by the clinical profiles of over 3000 breast cancer patients in the Swiss canton of Ticino, will also be contributed by the institute.
The evaluation will be done through the use of interviews, questionnaires and observation of the experts at work, with the objective of \textit{clinically validating} the system and \textit{assessing its capacity to relate to the medical professionals} in a meaningful manner.

The objective of such a methodology will be to both validate the claims made by \citet{Butz2018} and, more generally, to investigate the assertions made in the literature regarding the explainability of Bayesian networks (see Section \ref{sec:explainability-in-bayesian-networks}); at the same time, the aim is to also identify which characteristics of these models may be the most important in enabling their \textit{comprehensibility}.

### 2019-11-13 abstract.tex
Thesis complete

Our societies are delegating an ever increasing number of decisions to artificially intelligent systems and the need to understand the rationale for these is becoming progressively more apparent.
Especially in mission-critical domains as is the medical one, the demand for users to understand the \textit{why} of an automated decision is crucial.
%Unfortunately, the field of Explainable AI is itself confused about the exact properties it is trying to promote in machine learning systems and consequently of how to bring them about and evaluate them.

One of the main gaps in the current literature is the rarity of \textit{explainability} methods validated by real humans, in concrete settings.
This thesis aims to address this omission by focusing on assessing the explanatory powers of Bayesian Networks.
Such an evaluation will take place in the medical domain and will be done in collaboration with expert clinicians, employees of an institutional medical partner.
To this end, a proof of concept Bayesian Network-based system will be developed, applied to a real medical data set and evaluated in its clinical relevance and explanatory powers by the expert users.
The former will be tested by executing a series of clinical questions on the system and comparing the outcome with that expected by the experts while the latter shall be evaluated by a think-aloud study and by a questionnaire.

The developed tool has proved its clinical relevance and ability to meaningfully interact with expert medical users.
It is thus a step in the direction of validating the supposed explanatory powers of Bayesian Networks, even if not all the characteristics that were expected to be important in making these models more explainable than other machine learning techniques have been confirmed as such.

>>>>

Our societies are delegating an ever increasing number of decisions to artificially intelligent systems and the need to understand the rationale for these is becoming progressively more apparent.
Especially in mission-critical domains as is the medical one, the demand for users to understand the \textit{why} of an automated decision is crucial.
%Unfortunately, the field of explainable AI is itself confused about the exact properties it is trying to promote in machine learning systems and consequently of how to bring them about and evaluate them.

One of the main gaps in the current literature is the scarcity of \textit{explainability} methods validated by real humans, in concrete settings.
This thesis aims to address this omission by focusing on assessing the explanatory powers of Bayesian networks.
Such an evaluation takes place in the medical domain and is done in collaboration with expert clinicians, employees of an institutional medical partner.
To this end, a proof of concept Bayesian network-based system is developed, applied to a real medical data set and evaluated in its clinical relevance and explanatory powers by the expert users.
The former is tested by executing a series of clinical questions on the system and comparing the outcome with that expected by the experts while the latter is evaluated by a think-aloud study and by a questionnaire.

The developed tool has proved its clinical relevance and ability to meaningfully interact with expert medical users.
It is thus a step in the direction of validating the supposed explanatory powers of Bayesian networks, even if not all the characteristics that were expected to be important in making these models more explainable than other machine learning techniques, have been confirmed as such.

### 2019-11-13 annex1.tex
Thesis complete

\newcounter{annexcounter}

\begin{sidewaystable*}[h]
  \centering
  \captionsetup{name=Annex}
  \caption{Natural language questions that can be answered by conditional probability queries}
    \begin{tabularx}{\textwidth}{lXllllX}
	    \toprule
	    \textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\textbf{1} & At diagnosis, if estrogen receptors are negative, is tumor proliferative index high? & validation & ki67  & >30\% & estrogeni & 0-10\% \\
		\addlinespace
		\textbf{2} & At diagnosis, if estrogen receptors are negative, is the risk of metastases low? & validation & pM sub & pM=0  & estrogeni & 0-10\% \\
		\addlinespace
		\multirow{2}[0]{*}{\textbf{3}} & \multirow{2}[0]{4cm}{If estrogen receptors are negative and tumor proliferative index is  high at diagnosis, is the risk of metastases low?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{pM sub} & \multirow{2}[0]{*}{pM=0} & estrogeni & 0-10\% \\
		      &       &       &       &       & ki67  & >30\% \\
		\addlinespace[9ex]
	      \textbf{4} & If the diagnosis of mammary carcinoma happened at a young age, is tumour proliferative index high? & validation/research & ki67  & >30\% & eta arrotondata & <40 \\
      	\addlinespace
		\textbf{5} & If the histologic diagnosis is of lobular carcinoma, is the expression of the c-erbB2 marker absent? & validation & cerb  & 0 \& 1 & morfologia & Lobular carcinoma, NOS \\
		\addlinespace
		\textbf{6} & If the tumour is large, is lymph node involvement more probable? & validation & pN sub & pN!=0 & pT sub & pT>=2  \\  
		\addlinespace[2ex]
		\multirow{2}[0]{*}{\textbf{7}} & \multirow{2}[0]{4cm}{If the tumour is large and lymph nodes are involved, is the risk of metastases low at diagnosis?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{pM sub} & \multirow{2}[0]{*}{pM=0} & pT sub & pT>=2  \\ 
		&       &       &       &       & pN sub & pN!=0 \\  
\end{tabularx}
	\refstepcounter{annexcounter}
	\label{ann:conditionalprobability1}
\end{sidewaystable*}

\begin{sidewaystable*}[h]
  \centering
  \captionsetup{name=Annex}
  \caption{Natural language questions that can be answered by conditional probability queries}
    \begin{tabularx}{\textwidth}{lXllllX}
	    \toprule
	    \textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule      
		\textbf{8} & If the tumour is of high grade at diagnosis, is the risk of metastases low? & validation & pM sub & pM=0  & differenziazione & poco differenziato \\
		\multirow{2}[0]{*}{\textbf{9}} & \multirow{2}[0]{4cm}{In young patients, if estrogen receptors are negative, is tumor proliferative index high?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{ki67} & \multirow{2}[0]{*}{>30\%} & estrogeni & 0-10\% \\
		      &       &       &       &       & eta arrotondata & <40 \\	  
	    \addlinespace[6ex]
		\multirow{2}[0]{*}{\textbf{10}} & \multirow{2}[0]{4cm}{In young patients, if estrogen receptors are negative, is the risk of metastases low?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{pM sub} & \multirow{2}[0]{4cm}{pM=0} & estrogeni & 0-10\% \\
		      &       &       &       &       & eta arrotondata & <40 \\
      \addlinespace[6ex]
		\multirow{3}[0]{*}{\textbf{11}} & \multirow{3}[0]{4cm}{In young patients, if estrogen receptors are negative and tumor proliferative index is  high at diagnosis, is the risk of metastases low?} & \multirow{3}[0]{*}{validation} & \multirow{3}[0]{*}{pM sub} & \multirow{3}[0]{*}{pM=0} & estrogeni & 0-10\% \\
		      &       &       &       &       & ki67  & >30\% \\
		      &       &       &       &       & eta arrotondata & <40 \\
      \addlinespace[9ex]
	      \multirow{3}[0]{*}{\textbf{12}} & \multirow{3}[0]{4cm}{How does the tumoural grade change if I know the oestrogen expression?} & \multirow{3}[0]{*}{research} & \multirow{3}[0]{*}{differenziazione} &  ben differenziato & \multirow{3}[0]{*}{estrogeni} & negativi (0-10\%) \\
		      &       &       &       &  moderatamente differenziato &       & debolmente positivo (10-50\%) \\
		      &       &       &       &  poco differenziato &       & fortemente positivo (>50\%) \\
	      \addlinespace[1ex]
		\multirow{4}[0]{*}{\textbf{13}} & \multirow{4}[0]{4cm}{How does the oestrogen expression change if I know the proliferation index?} & \multirow{4}[0]{*}{research} & \multirow{4}[0]{*}{estrogeni} &  negativi (0-10\%) & \multirow{4}[0]{*}{ki67} & negativo (0-14\%) \\
		      &       &       &       &  debolmente positivi (10-50\%) &       & 14-20\% \\
		      &       &       &       & \multirow{2}[0]{*}{fortemente positivi (>50\%)} &       & 20-30\% \\
		      &       &       &       &       &       & positivi (>30\%) \\
      \addlinespace[1ex]
	      \textbf{14} & Does the negative expression of progestinic receptors influence lymph nodes' state? & validation & pN sub & pN!=0  & progestinici & 0-10\% \\
      \end{tabularx}
	\refstepcounter{annexcounter}
	\label{ann:conditionalprobability2}
\end{sidewaystable*}

\begin{sidewaystable}[h]
	\centering
	\captionsetup{name=Annex}
	\caption{Natural language questions that can be answered by d-separation queries}
	\begin{tabularx}{\textwidth}{lXllllX}
		\toprule
		\textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\textbf{14} & Which clinical-pathological variables influence the lymph nodes' state at diagnosis? & research & pN    & -     & -     & - \\
		\textbf{15} & Which clinical-pathological variables influence tumoural proliferation index? & research & ki67  & -     & -     & - \\
		\textbf{16} & Which clinical-pathological variables influence the expression of the c-ERBB2 marker? & research & cerb  & \multicolumn{1}{r}{} & -     & - \\
		\textbf{17} & Which clinical-pathological variables influence the oestrogen expression? & research & recettori estrogeni & \multicolumn{1}{r}{} & -     & - \\
		\textbf{18} & Which clinical-pathological variables influence the tumoural grade? & research & differenziazione & \multicolumn{1}{r}{} & -     & - \\
		\textbf{19} & Which clinical-pathological variables influence the presence of metastases at diagnosis? & research & pM    & \multicolumn{1}{r}{} & -     & - \\
		\textbf{20} & Which clinical-pathological variables influence the tumoural dimension? & research & pT    & \multicolumn{1}{r}{} & -     & - \\
		\textbf{21} & Which clinical-pathological variables influence the age of the tumour onset? & research & eta   & -     & -     & - \\
		\end{tabularx}
	\refstepcounter{annexcounter}
	\label{ann:dseparation}
\end{sidewaystable}

\begin{sidewaystable}[h]
	\centering
	\captionsetup{name=Annex}
	\caption{Natural language questions that can be answered by conditional probability queries or, at a higher level, by d-separation queries}
	\begin{tabularx}{\textwidth}{lXllllX}
		\toprule
		\textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\multirow{2}[0]{*}{\textbf{22}} & \multirow{2}[0]{4cm}{In young patients, does a negative expression of the progestinic receptors influence the lymph nodes' state?} & \multirow{2}[0]{*}{research} & \multirow{2}[0]{*}{pN sub} & \multirow{2}[0]{*}{pN!=0} & progestinici & 0-10\% \\
	      &       &       &       &       & eta   & <40 \\
	      \addlinespace[9ex]
		\multicolumn{1}{r}{\textbf{23}} & Does a negative expression of progestinic receptors influence the tumoural proliferation index? & research & ki67  & >30\% & progestinici & 0-10\% \\
		\addlinespace
		\multirow{2}[0]{*}{\textbf{24}} & \multirow{2}[0]{4cm}{In young patients, does a negative expression of the progestinic receptors influence the tumoural proliferation index?} & \multirow{2}[0]{*}{research} & \multirow{2}[0]{*}{ki67} & \multirow{2}[0]{*}{>30\%} & progestinici & 0-10\% \\
		      &       &       &       &       & eta   & <40 \\
		      \addlinespace[10ex]
		\multirow{3}[0]{*}{\textbf{25}} & \multirow{3}[0]{4cm}{Does a negative expression of progestinic receptors influence the expression of the c-ERBB2 marker?} & \multirow{3}[0]{*}{research} & \multirow{3}[0]{*}{cerb} & 0 \& 1 & \multirow{3}[0]{*}{progestinici} & \multirow{3}[0]{*}{0-10\%} \\
		      &       &       &       & 2     &       &  \\
		      &       &       &       & 3     &       &  \\
		      \addlinespace[4ex]
		\multirow{4}[0]{*}{\textbf{26}} & \multirow{4}[0]{4cm}{In young patients, does a negative expression of the progestinic receptors influence the expression of the c-ERBB2 marker?} & \multirow{4}[0]{*}{research} & \multirow{4}[0]{*}{cerb} & 0 \& 1 & \multirow{3}[0]{*}{progestinici} & \multirow{3}[0]{*}{0-10\%} \\
		      &       &       &       & 2     &       &  \\
		      &       &       &       & \multirow{2}[0]{*}{3} &       &  \\
		      &       &       &       &       & eta   & <40 \\
		\end{tabularx}
	\refstepcounter{annexcounter}
	\label{ann:conditionalanddseparation}
\end{sidewaystable}

\begin{sidewaystable}[h]
	\centering
	\captionsetup{name=Annex}
	\caption{Natural language questions that can be answered by MPE queries}
	\begin{tabularx}{\textwidth}{lXllllX}
		\toprule
		\textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\multirow{3}[0]{*}{\textbf{27}} & \multirow{3}[0]{4cm}{How are tumours characterised by a triple negative profile from the point of view of the other clinical-pathological variables?} & \multirow{3}[0]{*}{research} & \multirow{3}[0]{*}{-} & \multirow{3}[0]{*}{-} & cerb  & 0 \\
		\addlinespace[9ex]
	      &       &       &       &       & recettori estrogeni & negativo \\
	      &       &       &       &       & recettori progestinici & negativo \\
		\textbf{28} & How are tumours characterised by high ki67 from the point of view of the other clinical-pathological variables? & research & -     & -     & ki67  & >30 \\
		\addlinespace[2ex]
		\textbf{29} & How are tumours characterised by nodes involvement from the point of view of the other clinical-pathological variables? & research & -     & -     & pN    & !0 \\
		\end{tabularx}
	\refstepcounter{annexcounter}
	\label{ann:mpe}
\end{sidewaystable}

\begin{framed}
	\begin{center}
		{\huge Explainability evaluation questionnaire}
	\end{center}
	{\Large Confidence}
	\begin{enumerate} 
		\item Did the tool increase the confidence in diagnosis when diagnostic screening results were missing for a patient?  Why? \\
		O Yes O No
		\item Did the tool help in characterising a particular patient's profile? \\
		O Not at all O Somewhat O Absolutely
		\item Did the tool help in your confidence of understanding the cohort characteristics?  How? \\
		O Not at all O Somewhat O Absolutely
		\item Did the tool improve your confidence in your clinical decision-making?  How? \\ 
		O Not at all O Somewhat O Absolutely
		\item Did having the tool at your disposal improve your confidence when making time-constrained decisions?  How? (for example, did it improve confidence in prioritising some tests over others?) \\
		O Not at all O Somewhat O Absolutely
	\end{enumerate}
	{\Large Features}
	\begin{enumerate}[resume]
		\item Given the modes of interaction with the system labelled as \enquote{dialogues}, do you think you would have had more difficulty in interpreting the data without the these modalities? \\
		O No O Maybe O Yes
		\item Was natural language useful during the interaction?  Why? \\
		O No O Maybe O Yes
		\item Which type of \enquote{dialogue} did you feel was most useful? Why? \\
		O Exhaustive O Separations O Thresholded O A combination of the previous O All O None
		\item Did you feel that the dialogue helped you in cases of uncertainty?  If yes, how?  If no, why? \\
		O No O Somewhat O Yes
		\item Did you feel that the \enquote{dialogue} helped your clinical decision-making?  If yes, how?  If no, why? \\
		O No O Somewhat O Yes
		\item Did the generation of \enquote{counterfactual branches} help in your understanding of the data?  Why? \\
		O No O Somewhat O Yes
		\item Given the interaction mode labelled \enquote{Pseudo-MPE query}, how would you rate the solutions it proposed from a point of view of their understandability? (1 poor, 5 good) \\
		O 1 O 2 O 3 O 4 O 5
		\item How would you rate the Pseudo-MPE solutions from a point of view of their clinical usefulness? \\
		O 1 O 2 O 3 O 4 O 5
		\item Do you feel that the interaction mode labelled as \enquote{MPE query} gave better solutions than that labelled \enquote{Pseudo-MPE query}?  Why? \\
		O No O Maybe O Yes
		\item Did you find the \enquote{Pseudo-MPE} or \enquote{MPE} interaction mode the most useful?  Why? \\
		O Pseudo-MPE O MPE O Both O None
		\item How important was the highlighting of the independencies between variables? \\
		O 1 O 2 O 3 O 4 O 5
		\item Do you think you would have had more difficulty in interpreting the data without the correlation strength displayed? \\
		O No O Maybe O Yes
		\item Do you think you would have had more difficulty in interpreting the data without visualisations? \\
		O No O Maybe O Yes
		\item Do you think you would have had more difficulty in interpreting the data without natural language output? \\
		O No O Maybe O Yes
	\end{enumerate}
	{\Large Time}
	\begin{enumerate}[resume]
		\item How would you rate the time it took to understand the dialogues' outputs?  Which of the three was best? (1 bad, 5 good) \\
		O 1 O 2 O 3 O 4 O 5
		\item How would you rate the time it took to understand the conditional probability query's outputs \\
		O 1 O 2 O 3 O 4 O 5
		\item How would you rate the time it took to understand the MPE and Pseudo-MPE query's outputs? \\
		O 1 O 2 O 3 O 4 O 5
		\item Did natural language help in reducing the time needed to understand the outputs? \\
		O No O Somewhat O Yes
		\item Did visualisations help in reducing the time needed to understand the outputs? \\
		O No O Somewhat O Yes
	\end{enumerate}
	{\Large Tool}
	\begin{enumerate}[resume]
		\item Which interaction modes did you feel could be the most useful?  Why? \\
		O Plot model O Independencies O Conditional Probability Query O Pseudo-MPE and MPE O Dialogues
		\item Which interaction modes did you use the most?  Why? \\
		O Plot model O Independencies O Conditional Probability Query O Pseudo-MPE and MPE O Dialogues
		\item How did you use the tool in your day-to-day work?
		\item Is the tool missing any functionality that would address your needs?  If yes, which ones? \\
		O No O Yes
		\item Did you have any difficulties in understanding which functionalities to use to address your needs?  If yes, when? \\
		O No O Yes
		\item Did you have any difficulties in understanding the functionalities during usage?  If yes, when? \\
		O No O Yes
		\item If you answered Yes to the previous question, how do you think this could be addressed?
		\item Could you suggest any functionalities you would like to be implemented?
	\end{enumerate}
	{\Large Clinical}
	\begin{enumerate}[resume]
		\item Did the tool help in recovering missing features of patients thus supporting diagnostic profile creation and decision making? If yes, which is/are the feature/s that benefited the most? \\
		O No O Yes
		\item Did any of the tool's predictions have clinical confirmation later on?  If yes, how? \\
		O No O Yes
		\item Did the tool help in highlighting new relationships between variables? \\
		O No O Yes
		\item Did the tool help in highlighting new patient subgroups? \\
		O No O Yes
	\end{enumerate}
	{\Large Satisfaction}
	\begin{enumerate}[resume]
		\item What is your general satisfaction with the tool? For what reasons? \\
		O Completely dissatisfied O Somewhat dissatisfied O Neutral O Somewhat satisfied O Completely satisfied
	\end{enumerate}
	\refstepcounter{annexcounter}
	\label{ann:questionnaire}
\end{framed}

>>>>

\chapter{Natural Language Questions} \label{app:natural-language-questions}
%\captionsetup[table]{list=no}
%\newcounter{annex}
%\newcounter{subannex}[annex]
%\setcounter{annex}{1}
%\setcounter{subannex}{1}
%\renewcommand{\thetable}{\arabic{annex}.\arabic{subannex}}

This annex contains the natural language questions formulated by the medical partner - Istituto Cantonale di Patologia - with the objective of clinically validating the proof of concept system developed as part of the methods of this thesis.
These questions are referenced in the relevant Subsection \ref{subsec:clinical-validation-methodology}.

\begin{sidewaystable*}[h]
  \centering
  %\captionsetup{name=Appendix}
  \caption{Natural language questions answerable by conditional probability queries}
    \begin{tabularx}{\textwidth}{lXllllX}
	    \toprule
	    \textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\textbf{1} & At diagnosis, if estrogen receptors are negative, is tumor proliferative index high? & validation & ki67  & >30\% & estrogeni & 0-10\% \\
		\addlinespace
		\textbf{2} & At diagnosis, if estrogen receptors are negative, is the risk of metastases low? & validation & pM sub & pM=0  & estrogeni & 0-10\% \\
		\addlinespace
		\multirow{2}[0]{*}{\textbf{3}} & \multirow{2}[0]{4cm}{If estrogen receptors are negative and tumor proliferative index is  high at diagnosis, is the risk of metastases low?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{pM sub} & \multirow{2}[0]{*}{pM=0} & estrogeni & 0-10\% \\
		      &       &       &       &       & ki67  & >30\% \\
		\addlinespace[9ex]
	      \textbf{4} & If the diagnosis of mammary carcinoma happened at a young age, is tumour proliferative index high? & validation/research & ki67  & >30\% & eta arrotondata & <40 \\
      	\addlinespace
		\textbf{5} & If the histologic diagnosis is of lobular carcinoma, is the expression of the c-erbB2 marker absent? & validation & cerb  & 0 \& 1 & morfologia & Lobular carcinoma, NOS \\
		\addlinespace
		\textbf{6} & If the tumour is large, is lymph node involvement more probable? & validation & pN sub & pN!=0 & pT sub & pT>=2  \\  
		\addlinespace[2ex]
		\multirow{2}[0]{*}{\textbf{7}} & \multirow{2}[0]{4cm}{If the tumour is large and lymph nodes are involved, is the risk of metastases low at diagnosis?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{pM sub} & \multirow{2}[0]{*}{pM=0} & pT sub & pT>=2  \\ 
		&       &       &       &       & pN sub & pN!=0 \\  
\end{tabularx}
	%\refstepcounter{subannex}
	\label{app:conditionalprobability1}
\end{sidewaystable*}

\begin{sidewaystable*}[h]
  \centering
%  %\captionsetup{name=Appendix}
  \caption{Natural language questions answerable by conditional probability queries}
    \begin{tabularx}{\textwidth}{lXllllX}
	    \toprule
	    \textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule      
		\textbf{8} & If the tumour is of high grade at diagnosis, is the risk of metastases low? & validation & pM sub & pM=0  & differenziazione & poco differenziato \\
		\multirow{2}[0]{*}{\textbf{9}} & \multirow{2}[0]{4cm}{In young patients, if estrogen receptors are negative, is tumor proliferative index high?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{ki67} & \multirow{2}[0]{*}{>30\%} & estrogeni & 0-10\% \\
		      &       &       &       &       & eta arrotondata & <40 \\	  
	    \addlinespace[6ex]
		\multirow{2}[0]{*}{\textbf{10}} & \multirow{2}[0]{4cm}{In young patients, if estrogen receptors are negative, is the risk of metastases low?} & \multirow{2}[0]{*}{validation} & \multirow{2}[0]{*}{pM sub} & \multirow{2}[0]{4cm}{pM=0} & estrogeni & 0-10\% \\
		      &       &       &       &       & eta arrotondata & <40 \\
      \addlinespace[6ex]
		\multirow{3}[0]{*}{\textbf{11}} & \multirow{3}[0]{4cm}{In young patients, if estrogen receptors are negative and tumor proliferative index is  high at diagnosis, is the risk of metastases low?} & \multirow{3}[0]{*}{validation} & \multirow{3}[0]{*}{pM sub} & \multirow{3}[0]{*}{pM=0} & estrogeni & 0-10\% \\
		      &       &       &       &       & ki67  & >30\% \\
		      &       &       &       &       & eta arrotondata & <40 \\
      \addlinespace[9ex]
	      \multirow{3}[0]{*}{\textbf{12}} & \multirow{3}[0]{4cm}{How does the tumoural grade change if I know the oestrogen expression?} & \multirow{3}[0]{*}{research} & \multirow{3}[0]{*}{differenziazione} &  ben differenziato & \multirow{3}[0]{*}{estrogeni} & negativi (0-10\%) \\
		      &       &       &       &  moderatamente differenziato &       & debolmente positivo (10-50\%) \\
		      &       &       &       &  poco differenziato &       & fortemente positivo (>50\%) \\
	      \addlinespace[1ex]
		\multirow{4}[0]{*}{\textbf{13}} & \multirow{4}[0]{4cm}{How does the oestrogen expression change if I know the proliferation index?} & \multirow{4}[0]{*}{research} & \multirow{4}[0]{*}{estrogeni} &  negativi (0-10\%) & \multirow{4}[0]{*}{ki67} & negativo (0-14\%) \\
		      &       &       &       &  debolmente positivi (10-50\%) &       & 14-20\% \\
		      &       &       &       & \multirow{2}[0]{*}{fortemente positivi (>50\%)} &       & 20-30\% \\
		      &       &       &       &       &       & positivi (>30\%) \\
      \addlinespace[1ex]
	      \textbf{14} & Does the negative expression of progestinic receptors influence lymph nodes' state? & validation & pN sub & pN!=0  & progestinici & 0-10\% \\
      \end{tabularx}
	%\refstepcounter{subannex}
	\label{app:conditionalprobability2}
\end{sidewaystable*}

\begin{sidewaystable}[h]
	\centering
	%\captionsetup{name=Appendix}
	\caption{Natural language questions answerable by d-separation queries}
	\begin{tabularx}{\textwidth}{lXllllX}
		\toprule
		\textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\textbf{14} & Which clinical-pathological variables influence the lymph nodes' state at diagnosis? & research & pN    & -     & -     & - \\
		\textbf{15} & Which clinical-pathological variables influence tumoural proliferation index? & research & ki67  & -     & -     & - \\
		\textbf{16} & Which clinical-pathological variables influence the expression of the c-ERBB2 marker? & research & cerb  & \multicolumn{1}{r}{} & -     & - \\
		\textbf{17} & Which clinical-pathological variables influence the oestrogen expression? & research & recettori estrogeni & \multicolumn{1}{r}{} & -     & - \\
		\textbf{18} & Which clinical-pathological variables influence the tumoural grade? & research & differenziazione & \multicolumn{1}{r}{} & -     & - \\
		\textbf{19} & Which clinical-pathological variables influence the presence of metastases at diagnosis? & research & pM    & \multicolumn{1}{r}{} & -     & - \\
		\textbf{20} & Which clinical-pathological variables influence the tumoural dimension? & research & pT    & \multicolumn{1}{r}{} & -     & - \\
		\textbf{21} & Which clinical-pathological variables influence the age of the tumour onset? & research & eta   & -     & -     & - \\
		\end{tabularx}
	%\refstepcounter{subannex}
	\label{app:dseparation}
\end{sidewaystable}

\begin{sidewaystable}[h]
	\centering
	%\captionsetup{name=Appendix}
	\caption{Natural language questions answerable by conditional probability queries or, at a higher level, by d-separation queries}
	\begin{tabularx}{\textwidth}{lXllllX}
		\toprule
		\textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\multirow{2}[0]{*}{\textbf{22}} & \multirow{2}[0]{4cm}{In young patients, does a negative expression of the progestinic receptors influence the lymph nodes' state?} & \multirow{2}[0]{*}{research} & \multirow{2}[0]{*}{pN sub} & \multirow{2}[0]{*}{pN!=0} & progestinici & 0-10\% \\
	      &       &       &       &       & eta   & <40 \\
	      \addlinespace[9ex]
		\multicolumn{1}{r}{\textbf{23}} & Does a negative expression of progestinic receptors influence the tumoural proliferation index? & research & ki67  & >30\% & progestinici & 0-10\% \\
		\addlinespace
		\multirow{2}[0]{*}{\textbf{24}} & \multirow{2}[0]{4cm}{In young patients, does a negative expression of the progestinic receptors influence the tumoural proliferation index?} & \multirow{2}[0]{*}{research} & \multirow{2}[0]{*}{ki67} & \multirow{2}[0]{*}{>30\%} & progestinici & 0-10\% \\
		      &       &       &       &       & eta   & <40 \\
		      \addlinespace[10ex]
		\multirow{3}[0]{*}{\textbf{25}} & \multirow{3}[0]{4cm}{Does a negative expression of progestinic receptors influence the expression of the c-ERBB2 marker?} & \multirow{3}[0]{*}{research} & \multirow{3}[0]{*}{cerb} & 0 \& 1 & \multirow{3}[0]{*}{progestinici} & \multirow{3}[0]{*}{0-10\%} \\
		      &       &       &       & 2     &       &  \\
		      &       &       &       & 3     &       &  \\
		      \addlinespace[4ex]
		\multirow{4}[0]{*}{\textbf{26}} & \multirow{4}[0]{4cm}{In young patients, does a negative expression of the progestinic receptors influence the expression of the c-ERBB2 marker?} & \multirow{4}[0]{*}{research} & \multirow{4}[0]{*}{cerb} & 0 \& 1 & \multirow{3}[0]{*}{progestinici} & \multirow{3}[0]{*}{0-10\%} \\
		      &       &       &       & 2     &       &  \\
		      &       &       &       & \multirow{2}[0]{*}{3} &       &  \\
		      &       &       &       &       & eta   & <40 \\
		\end{tabularx}
	%\refstepcounter{subannex}
	\label{app:conditionalanddseparation}
\end{sidewaystable}

\begin{sidewaystable}[h]
	\centering
	%\captionsetup{name=Appendix}
	\caption{Natural language questions answerable by MPE queries}
	\begin{tabularx}{\textwidth}{lXllllX}
		\toprule
		\textbf{\#} & Natural language question & Type  & Target variable & Target value & Evidence variable & Evidence value \\
		\midrule
		\multirow{3}[0]{*}{\textbf{27}} & \multirow{3}[0]{4cm}{How are tumours characterised by a triple negative profile from the point of view of the other clinical-pathological variables?} & \multirow{3}[0]{*}{research} & \multirow{3}[0]{*}{-} & \multirow{3}[0]{*}{-} & cerb  & 0 \\
		\addlinespace[9ex]
	      &       &       &       &       & recettori estrogeni & negativo \\
	      &       &       &       &       & recettori progestinici & negativo \\
		\textbf{28} & How are tumours characterised by high ki67 from the point of view of the other clinical-pathological variables? & research & -     & -     & ki67  & >30 \\
		\addlinespace[2ex]
		\textbf{29} & How are tumours characterised by nodes involvement from the point of view of the other clinical-pathological variables? & research & -     & -     & pN    & !0 \\
		\end{tabularx}
	%\refstepcounter{subannex}
	\label{app:mpe}
\end{sidewaystable}
